---
title: "Define Custom Response Distributions"
author: "Allen Baumgardner-Zuzik"
date: "February 3, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(brms)
```

Author: Paul Burkner
Date: 2020-11-02

## Introduction

The **brms package** comes with a lot of built-in response distributions – usually called *families* in R – to specify among others linear, count data, survival, response times, or ordinal models (see `help(brmsfamily`) for an overview). Despite supporting over two dozen families, there is still a long list of distributions, which are not natively supported. The present vignette will explain how to specify such custom families in **brms**. By doing that, users can benefit from the modeling flexibility and post-processing options of **brms** even when using self-defined response distributions.

## A Case Study

As a case study, we will use the `cbpp` data of the **lme4 package**, which describes the development of the CBPP disease of cattle in Africa. The data set contains four variables: `period` (the time period), `herd` (a factor identifying the cattle herd), `incidence` (number of new disease cases for a given herd and time period), as well as `size` (the herd size at the beginning of a given time period).

```{r, cache=TRUE}
data("cbpp", package = "lme4")
head(cbpp)
```

In a first step, we will be predicting `incidence` using a simple binomial model, which will serve as our baseline model. For observed number of events y (`incidence` in our case) and total number of trials $T$(`size` ), the probability mass function of the binomial distribution is defined as

$$P(y|T,p)=\binom{T}{y}p^{y}(1−p)^{N−y}$$

where $p$ is the event probability. In the classical binomial model, we will directly predict $p$ on the logit-scale, which means that for each observation i we compute the success probability $p_i$ as

$$p_{i}=\frac{exp(η_{i})}{1+exp(η_{i})}$$

where $η_i$ is the linear predictor term of observation $i$ (see `vignette("brms_overview")` for more details on linear predictors in **brms**). Predicting `incidence` by `period` and a varying intercept of `herd` is straight forward in **brms**:


```{r, cache=TRUE, results='hide'}
fit1 <- brm(incidence | trials(size) ~ period + (1|herd),
           data = cbpp, family = binomial())
```

In the summary output, we see that the incidence probability varies substantially over herds, but reduces over the course of the time as indicated by the negative coefficients of `period`.

```{r}
summary(fit1)
```

A drawback of the binomial model is that – after taking into account the linear predictor – its variance is fixed to $Var(y_{i})=T_{i}p_{i}(1−p_{i})$. All variance exceeding this value cannot be not taken into account by the model. There are multiple ways of dealing with this so called *overdispersion* and the solution described below will serve as an illustrative example of how to define custom families in **brms**.
